---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(teethr)
library(dplyr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# teethr

<!-- badges: start -->
[![R-CMD-check](https://github.com/bbartholdy/teethr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bbartholdy/teethr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of teethr (pronounced teether) is to provide a set of tools to calculate
dental indices, such as caries ratio, and calculus index, and facilitate
visualisation of data related to dental diseases.

## Installation

You can install the development version of teethr like so:

``` r
devtools::install_github("bbartholdy/teethr")
```

## Example

First, dental data are combined with demographic data and converted to long format with
`dental_longer()`. `dental_longer()` also adds information on tooth type and
position that can be used to calculate tooth- and position-specific indices.

```{r example, eval=FALSE}
library(teethr)
library(dplyr)

mb11_calculus_long <- mb11_calculus %>% 
  dental_longer(-id, names_sep = "_") %>% # make longer all columns except 'id'
  dental_join() # add tooth information (position, type, etc.)
  inner_join(mb11_demography, by = "id") # combine with age and sex
mb11_calculus_long
```

```{r example-out, echo=FALSE}
mb11_calculus_long <- mb11_calculus %>% 
  dental_longer(-id, names_sep = "_") %>% # make longer all columns except 'id'
  inner_join(mb11_demography, by = "id") # combine with age and sex
mb11_calculus_long
```


The long-format data can be used to calculate indices, such as the dental calculus
index using the `calculus_index()` function. The function also provides the number
of surfaces scored, number of teeth, and sum of scores for verification. The
original method recommends grouping the dentition by quadrant to obtain four
indices. This can be done for the whole sample,

```{r calculus-index}
mb11_calculus_long %>%
  group_by(quadrant) %>% 
  calculus_index()
```

or by sex,

```{r calc-demography}
mb11_calculus_long %>%
  # groups can also be added directly to calculus_index()
  calculus_index(sex, quadrant) %>% 
  select(sex, quadrant, index) # remove unnecessary outputs
```

or by age and sex.

```{r calc-age-sex}
mb11_calculus_long %>%
  calculus_index(sex, age, quadrant) %>% 
  select(sex, quadrant, index) # remove unnecessary outputs
```


```{r calc-plot}
library(ggplot2)
mb11_calculus_long %>% 
  calculus_index(id, age, quadrant) %>% 
  #filter(quadrant == "UP") %>% 
  ggplot(aes(x = age, y = index, fill = age)) +
    #geom_violin(aes(fill = sex)) +
    geom_boxplot(width = 0.5) +
    scale_fill_viridis_d() +
    facet_wrap(~ quadrant) +
    theme_bw()
```

Dental caries ratios can be calculated using the `caries_ratio` function.

```{r caries-ratio}
mb11_caries_long <- mb11_caries %>% 
  dental_longer(-id) %>%
  dental_join()

mb11_caries_long %>% 
  count_caries(caries = score, no_lesion = "none") %>% # convert location to lesion count
  group_by(type) %>% 
  dental_ratio(count = caries_count)
```


The package also facilitates working with the {tidyverse}.

```{r}
# count number of teeth per individual
mb11_caries_long %>% 
  remove_missing(vars = "score") %>% 
  count(id)
```

